Time--Frequency Scattering Accurately Models Auditory Similarities Between Instrumental Playing Techniques

### 2020-07-21 paper/paper.tex
fix 2 factor in technique-similarity
comment on line 936
final comments

Timbre involves multiple time scales in conjunction, from a few microseconds for an attack transient to several seconds for a sustained tone.
Therefore, computational models of timbre perception must summarize acoustic information over a long analysis window, typically amounting to $10^{4}$ digital audio samples or more \cite{joder2009taslp}.
Mapping this input to a feature space in which distances denote timbral dissimilarity requires a data-driven stage of dimensionality reduction.
In this respect, the scientific literature exhibits a methodological divide as regards the collection of human-annotated data \cite{siedenburg2016jnmr}: while the field of MIR mostly encodes timbre under the form of ``audio tags'', music psychology mostly measures timbre similarity directly from pairwise \ml{similarity} judgments.

@@

\subsection*{Inter-technique similarity}
From the consensus clustering graph $\mathcal{G}_0$, we also derive a technique-wise similarity matrix $\mathbf{A}_{\mathcal{T}}$ whose rows and columns are defined on the finite set $\mathcal{T}$ of playing techniques.
As for $\mathbf{A}_{\mathcal{I}}$, we define the similarity between any two playing techniques $u$ and $v$ in $\mathcal{T}$ as the following ratio of set cardinalities:
\begin{equation}
\mathbf{A}_{\mathcal{T}}(u,v) = \dfrac{
2\,\Card \big\{ (m, n) \,\vert\, \mathrm{Technique}(m)=u ; \mathrm{Technique}(n)=v ; m \overset{\mathcal{G}_0}{\sim} n \big\}
}{
\Card \big\{n \,\vert\, \mathrm{Technique}(n) \in \{ u, v \} \big\}
}.
\label{eq:technique-similarity}
\end{equation}
\ml{why 2 ?}
Again, we run Ward's method to cluster playing techniques in $\mathcal{T}$ according to their similarity $\mathbf{A}_{\mathcal{T}}$.
Figure \ref{fig:consensusVsPt} displays the rearranged similarity matrix $\widetilde{\mathbf{A}}_{\mathcal{T}} : (u,v)\in \mathcal{T}^2 \mapsto \mathbf{A}_{\mathcal{T}}(\sigma_{\mathcal{T}}^{-1}(u), \sigma_{\mathcal{T}}^{-1}(v))$.

@@

Given an audio stimulus $\boldsymbol{x}_n$ from our perceptual study, we seek its position in the cluster graph $\mathcal{G}_0$.
Then, we identify its instrument--mute--technique (IMT) triplet, scrape for audio samples in SOL matching this triplet, and assign them all to the same cluster as the original audio stimulus.
We repeat the same procedure for all $N=78$ nodes in $\mathcal{G}_0$, resulting in $N^\prime = 9346$ samples in total.
This is a form of semi-supervised label propagation: from a limited amount of human annotation in order to curate a relatively large subset of SOL, amounting to about one third of the entire dataset ($9346$ out of $25444$ samples).

@@

That being said, the above model is general enough to encompass a wide variety of recordings, not just instrumental sounds.
Indeed, we have strong assumptions that the tools used (scattering transforms and LMNN weighting matrices) do not depend on the type of sound being processed. \ml{Future work will demonstrate that they work for more general classes of sounds}.
To extend the model, it is only necessary to retrain the LMNN weighting matrix by supplying it with new cluster assignments.
These can again be obtained by performing a new clustering experiment with one or more human subjects.

>>>>

Timbre involves multiple time scales in conjunction, from a few microseconds for an attack transient to several seconds for a sustained tone.
Therefore, computational models of timbre perception must summarize acoustic information over a long analysis window, typically amounting to $10^{4}$ digital audio samples or more \cite{joder2009taslp}.
Mapping this input to a feature space in which distances denote timbral dissimilarity requires a data-driven stage of dimensionality reduction.
In this respect, the scientific literature exhibits a methodological divide as regards the collection of human-annotated data \cite{siedenburg2016jnmr}: while the field of MIR mostly encodes timbre under the form of ``audio tags'', music psychology mostly measures timbre similarity directly from pairwise similarity judgments.

@@

\subsection*{Inter-technique similarity}
From the consensus clustering graph $\mathcal{G}_0$, we also derive a technique-wise similarity matrix $\mathbf{A}_{\mathcal{T}}$ whose rows and columns are defined on the finite set $\mathcal{T}$ of playing techniques.
As for $\mathbf{A}_{\mathcal{I}}$, we define the similarity between any two playing techniques $u$ and $v$ in $\mathcal{T}$ as the following ratio of set cardinalities:
\begin{equation}
\mathbf{A}_{\mathcal{T}}(u,v) = \dfrac{
\,\Card \big\{ (m, n) \,\vert\, \mathrm{Technique}(m)=u ; \mathrm{Technique}(n)=v ; m \overset{\mathcal{G}_0}{\sim} n \big\}
}{
\Card \big\{n \,\vert\, \mathrm{Technique}(n) \in \{ u, v \} \big\}
}.
\label{eq:technique-similarity}
\end{equation}
%\ml{why 2 ?} \vl: fixed
Again, we run Ward's method to cluster playing techniques in $\mathcal{T}$ according to their similarity $\mathbf{A}_{\mathcal{T}}$.
Figure \ref{fig:consensusVsPt} displays the rearranged similarity matrix $\widetilde{\mathbf{A}}_{\mathcal{T}} : (u,v)\in \mathcal{T}^2 \mapsto \mathbf{A}_{\mathcal{T}}(\sigma_{\mathcal{T}}^{-1}(u), \sigma_{\mathcal{T}}^{-1}(v))$.

@@

Given an audio stimulus $\boldsymbol{x}_n$ from our perceptual study, we seek its position in the cluster graph $\mathcal{G}_0$.
Then, we identify its instrument--mute--technique (IMT) triplet, scrape for audio samples in SOL matching this triplet, and assign them all to the same cluster as the original audio stimulus.
We repeat the same procedure for all $N=78$ nodes in $\mathcal{G}_0$, resulting in $N^\prime = 9346$ samples in total.
This is a form of semi-supervised label propagation: from a limited amount of human annotation, we curate a relatively large subset of SOL, amounting to about one third of the entire dataset ($9346$ out of $25444$ samples). % \vl: ?

@@

That being said, the above model is general enough to encompass a wide variety of recordings, not just instrumental sounds.
Indeed, we have strong assumptions that the tools used (scattering transforms and LMNN weighting matrices) do not depend strongly on the type of sound being processed. Future work will investigate whether more general classes of sounds are also well modeled.
To extend the model, it is only necessary to retrain the LMNN weighting matrix by supplying it with new cluster assignments.
These can again be obtained by performing a new clustering experiment with one or more human subjects.
